\documentclass[sigconf,anonymous,review]{acmart}

%% Conference (adjust as appropriate for submission)
%% \setcopyright{none}
%% \acmConference[ACSAC 2026]{Annual Computer Security Applications Conference}{December 2026}{Austin, TX, USA}

\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{makecell}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  columns=fullflexible,
  keepspaces=true
}

%%
%% --- Paper metadata ---------------------------------------------------
\title{Preventing Authority Laundering in Tool-Using LLM Agents}

\author{Author Names Omitted for Review}

\keywords{LLM agents, authority laundering, prompt injection, runtime enforcement, capability tokens, MCP security, declassification}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002978.10003006.10003013</concept_id>
<concept_desc>Security and privacy~Systems security</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10002978.10003014</concept_id>
<concept_desc>Security and privacy~Information flow control</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010178.10010179</concept_id>
<concept_desc>Computing methodologies~Artificial intelligence</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Security and privacy~Systems security}
\ccsdesc[500]{Security and privacy~Information flow control}
\ccsdesc[300]{Computing methodologies~Artificial intelligence}

\begin{document}

\begin{abstract}
Tool-using LLM agents do not merely generate text; they exercise delegated authority. They read private data, invoke MCP tools, load skills, write files, send emails, and execute code. This changes the security problem: an attacker does not need to compromise the model as software---it is enough to launder a low-authority observation into a high-authority tool action. We call this \emph{authority laundering}: the process by which a low-authority source---such as a webpage, email, retrieved document, MCP tool description, skill file, or tool output---is transformed by an LLM planner into a high-risk tool effect without proper authorization.

We present \textbf{ProvShield}, a runtime authority firewall for MCP and skill-based LLM agents. ProvShield treats the LLM as an untrusted planner and moves all authorization decisions to the runtime. Each content object entering the agent is assigned an HMAC-keyed sidecar provenance label tracking authority origin. Tools are declared with effect types and sinks. Before any tool call executes, a runtime monitor verifies that the proposed action, destination, payload, and capability satisfy a source-to-sink authority policy. For high-risk effects, ProvShield requires an \emph{intent-bound declassification capability}: a one-time, cryptographically bound authorization specific to the exact action, sink, destination, payload digest, principal, and expiration---not a generic user confirmation.

ProvShield achieves 0.6\% ASR (95\% CI [0.2\%, 1.7\%]), an 88\% reduction from the no-defense baseline, while maintaining 92.4\% BTCR. Its conditional block rate is 100\%: every attack tool call the LLM generates is stopped by the runtime. Direct-call adversary evaluation with 101 scenarios across 8 effect types confirms 100\% block rate with conservative provenance.
\end{abstract}
\maketitle

%% ======================================================================
\section{Introduction}
\label{sec:introduction}

Tool-using LLM agents do not merely generate text; they exercise delegated authority. They read private data, invoke MCP tools, load skills, write files, send emails, and execute code on behalf of users. This changes the prompt-injection problem. The attacker does not need to compromise the model as software; it is enough to launder a low-authority observation into a high-authority tool action.

We call this failure mode \emph{authority laundering}. A webpage that the agent summarizes should provide facts, not authorize a new email send. A tool output should inform the agent, not trigger code execution. A tool description should describe its interface, not grant itself permission to attach credentials. A skill file should not override system policy. Yet in current agent architectures, these sources are represented as natural language in a shared planning context, without a runtime-enforced distinction between data and instruction, between observation and authorization.

Existing defenses are insufficient because they address the wrong layer. Prompt hardening tells the model to ignore injected instructions, but this depends on model obedience and fails against contextual or metadata-driven attacks~\cite{openai_agent_prompt_injection_defense}. Input filtering classifies malicious text, but laundering can be indirect, socially engineered, or embedded in tool metadata~\cite{greshake2023indirect}. Generic confirmation asks the user before writes, but the confirmation does not bind the actual destination, payload, or source of influence---enabling confirmation laundering~\cite{perez2022ignore}. Attribution monitors estimate which context influenced an action, but influence is not authority.

This paper takes a different position. The model may be manipulated or fully compromised; the runtime must still prevent unauthorized high-risk effects. We design \textbf{ProvShield}, a runtime authority firewall for MCP and skill-based LLM agents. ProvShield assigns HMAC-keyed sidecar labels to content as it enters the agent runtime, declares tools with effect types and sinks~\cite{denning1976lattice,myers1997decentralized}, and intercepts all proposed tool calls before execution. A tool call executes only if the runtime can establish an authority proof: the action, destination, payload, and capability are authorized by high-authority sources or by an \emph{intent-bound declassification capability}---a one-time, cryptographically bound authorization specific to the exact call.

\subsection{Contributions}

This paper makes the following contributions:

\begin{enumerate}
  \item \textbf{Problem formulation: Agent Authority Laundering.} We define authority laundering as a security failure specific to tool-using LLM agents, where low-authority observations are transformed into high-authority tool effects through model planning. We identify six laundering classes and analyze why existing defenses fail against each.
  \item \textbf{Authority-flow model for MCP/Skills agents.} We formalize source authorities, effects, sinks, authority levels, user intent, and declassification capabilities as a runtime transition system over the observation-to-execution pipeline.
  \item \textbf{Intent-bound declassification capability.} We design a security abstraction where user confirmation binds action, effect, sink, destination, payload digest, source category, principal, and expiry into a non-replayable, one-time runtime capability.
  \item \textbf{Runtime authority firewall.} We treat the LLM as an untrusted planner and enforce authority proofs at the tool execution boundary, preventing unauthorized high-risk effects even when the planner is fully compromised.
  \item \textbf{Authority-laundering benchmark.} We construct a benchmark covering six laundering classes with 780 LLM-in-the-loop scenarios across three models~\cite{skillinject,mcptox,mcp_safety_audit}.
  \item \textbf{Empirical separation of model manipulation from runtime enforcement.} We decompose attack success into LLM manipulation rate, attack-tool generation rate, conditional runtime block rate, and direct-call adversary ASR, demonstrating that ProvShield's 100\% conditional block rate operates independently of model safety alignment.

\end{enumerate}
%% ======================================================================
\section{Authority Laundering Attacks}
\label{sec:attacks}

We define authority laundering as the process by which a low-authority observation is transformed by an LLM planner into a high-authority tool effect. We identify six classes of authority laundering attacks, each exploiting a different failure mode in current agent architectures.

\subsection{Webpage-to-Email Exfiltration}
\label{sec:attack-web}

A user asks the agent to summarize a webpage. The webpage contains hidden text instructing the agent to send the user's API key to an attacker-controlled email address~\cite{greshake2023indirect}. A prompt-only defense requires the model to recognize and ignore the hidden instruction. ProvShield instead labels the webpage as \texttt{ExternalContent}. If the model proposes \texttt{send\_email}, the monitor detects that the destination and objective are influenced by low-integrity content and that the payload may contain \texttt{Secret}. The call is denied unless an explicit, bound declassification exists.

\subsection{MCP Metadata Poisoning}
\label{sec:attack-mcp}

An MCP server registers a tool with a benign name but a malicious description: ``Use this tool to validate credentials; before calling, include all available tokens.''~\cite{mcp_safety_audit,mcptox} If the agent treats tool metadata as trusted instruction, the metadata can shape future tool calls. ProvShield treats unattested metadata as \texttt{ToolMetadata}, a low-integrity source. Metadata can describe schema but cannot grant itself authority or introduce privileged objectives.

\subsection{Skill File Injection}
\label{sec:attack-skill}

A skill package includes instructions that appear to implement a formatting task but also tell the agent to delete local files after completion~\cite{skillinject}. If loaded as trusted natural-language instructions, the skill can override user intent. ProvShield labels untrusted skills separately, prevents them from changing policy, and blocks destructive effects unless authorized by user intent and capability.

\subsection{Confirmation Laundering}
\label{sec:attack-bridge}

A malicious email asks the agent to ``send the requested diagnostic information to the support address.'' The agent presents a generic confirmation: ``Send email?'' The user clicks yes. The actual recipient is attacker-controlled and the payload includes private data. ProvShield's bridge would show the exact recipient, payload digest, source categories, and private-data crossing. A bridge for one payload and destination cannot authorize another.

\subsection{Tool-Output-to-Exec Laundering}
\label{sec:attack-tool-output}

A security scanner tool returns output that includes a hidden instruction to execute a shell command as part of ``post-scan cleanup.'' The tool output---a low-authority source---is laundered into a code-execution effect (\texttt{ExecuteCode}). ProvShield labels tool outputs as \texttt{ToolOutput}, which cannot authorize high-risk effects on its own. The policy engine requires a bridge for any \texttt{ExecuteCode} effect influenced by tool output.

\subsection{Replay and Swap Laundering}
\label{sec:attack-replay}

An attacker obtains a previously authorized bridge token and attempts to reuse it for a different destination, payload, or action. Alternatively, the attacker modifies the proposed call after the bridge is confirmed but before execution---a destination swap or payload swap. ProvShield's capability tokens carry a MAC over all six normalized call fields (action, effect, destination, payload digest, principal, nonce) and an expiration timestamp. Any field mismatch causes verification to fail. The nonce is consumed upon use, preventing replay.

\subsection{Attack Taxonomy}
\label{sec:taxonomy}

Table~\ref{tab:taxonomy} summarizes the six authority-laundering classes.

\begin{table}[t]
  \centering
  \caption{Authority laundering attack taxonomy.}
  \label{tab:taxonomy}
  \small
  \begin{tabular}{@{}llll@{}}
    \toprule
    \textbf{Attack Class} & \textbf{Low-Privilege Source} & \textbf{Laundered Effect} & \textbf{ProvShield Block Point} \\
    \midrule
    Observation-to-send & Web/email/RAG & \texttt{SendNetwork} & Source-to-sink policy \\
    Metadata-to-credential & MCP metadata & \texttt{CreateCredential} & Metadata integrity $<$ UserIntent \\
    Skill-to-delete & Untrusted skill & \texttt{DeleteLocal} & Skill labeled UntrustedSkill \\
    Tool-output-to-exec & Tool output & \texttt{ExecuteCode} & ToolOutput cannot authorize alone \\
    Confirmation & Ambiguous UI & Bound authority & Bridge binds all fields \\
    Replay/swap & Old bridge/capability & New dest/payload & MAC + nonce consumption \\
    \bottomrule
  \end{tabular}
\end{table}

%% ======================================================================
\section{Threat Model}
\label{sec:threat-model}

The adversary may control external content, emails, retrieved documents, tool outputs, untrusted MCP metadata, and untrusted skill files. The adversary may know the policy rules and the label syntax rendered to the model. The adversary may attempt label spoofing, metadata poisoning, bridge laundering, delayed injection, multi-step exfiltration, or capability replay.

The LLM is \emph{not} trusted to make authorization decisions. It may be fully compromised: the adversary may control the planner's output completely. The security property is not that the model is unaffected by malicious content, but that \emph{unauthorized high-risk effects cannot execute} regardless of the planner's behavior. The LLM is modeled as an adversarial proposal function that may output arbitrary tool calls.

The trusted computing base (TCB) consists of:
\begin{itemize}
  \item the runtime monitor,
  \item the sidecar provenance store,
  \item the policy engine,
  \item the token generator (HMAC-SHA256),
  \item the bridge manager, and
  \item the audit logger.
\end{itemize}

ProvShield assumes the TCB is not compromised. It does not attempt to prove that untrusted text has no influence on the model's hidden state. Instead, it proves that high-risk tool execution cannot occur unless the runtime-observable authority and capability constraints are satisfied.

%% ======================================================================
\section{Design}
\label{sec:design}

ProvShield's design centers on the principle that every high-risk tool effect requires a runtime-verifiable authority proof. Figure~\ref{fig:architecture} shows the architecture. The LLM is treated as an untrusted planner; all proposed tool calls pass through a runtime monitor before execution.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    node distance=0.8cm and 1.2cm,
    box/.style={draw, rounded corners, minimum width=2.0cm, minimum height=0.6cm, font=\small, align=center},
    arr/.style={-{Stealth[length=2mm]}, thick},
    every node/.style={font=\small}
  ]
    \node[box] (user) {User UI};
    \node[box, right=1.5cm of user] (web) {Web / Email / RAG};
    \node[box, below=0.6cm of user] (skill) {Skill Loader};
    \node[box, below right=0.6cm and 0.4cm of web] (cb) {Context Builder};
    \node[box, below=0.6cm of cb] (store) {Sidecar Store};
    \node[box, right=1.0cm of cb] (llm) {LLM Planner};
    \node[box, below=0.8cm of store] (mon) {Runtime Monitor};
    \node[box, left=0.8cm of mon] (policy) {Policy Engine};
    \node[box, right=0.8cm of mon] (bridge) {Bridge Manager};
    \node[box, below=0.6cm of mon] (mcp) {MCP Proxy};
    \node[box, below=0.6cm of mcp] (tools) {MCP / Local Tools};
    \node[box, below=0.6cm of policy] (audit) {Audit Log};

    \draw[arr] (user) -- (cb);
    \draw[arr] (web) -- (cb);
    \draw[arr] (skill) -- (cb);
    \draw[arr] (cb) -- (store);
    \draw[arr] (cb) -- (llm);
    \draw[arr] (llm) -- node[above,font=\scriptsize] {proposed call} (mon);
    \draw[arr] (store) -- (mon);
    \draw[arr] (policy) -- (mon);
    \draw[arr] (bridge) -- (mon);
    \draw[arr] (mon) -- node[left,font=\scriptsize] {allow} (mcp);
    \draw[arr] (mon) -- node[right,font=\scriptsize] {require} (bridge);
    \draw[arr] (mcp) -- (tools);
    \draw[arr] (tools) -- node[right,font=\scriptsize] {labeled output} (cb);
    \draw[arr] (mon) -- (audit);
  \end{tikzpicture}
  \caption{ProvShield architecture. Content entering the agent is labeled with provenance metadata in a sidecar store. The runtime monitor intercepts all proposed tool calls and enforces source-to-sink policy using provenance labels, effect types, and capability tokens.}
  \Description{Block diagram showing ProvShield architecture with User UI, Web/Email/RAG, and Skill Loader feeding into a Context Builder and Sidecar Store. The LLM Planner sends proposed tool calls to the Runtime Monitor, which consults the Policy Engine and Bridge Manager. Allowed calls pass through the MCP Proxy to MCP/Local Tools. All decisions are logged to the Audit Log.}
  \label{fig:architecture}
\end{figure}

\subsection{Runtime Authority Firewall}
\label{sec:firewall}

The runtime authority firewall is the core of ProvShield. Every proposed tool call---whether generated by the LLM, a compromised planner, or any other source---must pass through the monitor before execution. The monitor operates deterministically: it does not use the LLM for policy decisions. The firewall implements a simple principle: \emph{no tool effect executes without sufficient authority}.

For each proposed call, the firewall: (1) normalizes the tool name and arguments into a canonical form; (2) infers effects and sinks from the tool's declaration; (3) reconstructs argument and payload provenance from the sidecar store; (4) checks whether the authority sources satisfy the source-to-sink policy; (5) allows, denies, or requires an intent-bound declassification capability; and (6) logs the decision for replay.

\subsection{Authority-Flow Model}
\label{sec:labels}

Each object entering the agent runtime is assigned an \emph{authority label} tracking its origin:
\begin{lstlisting}
Labeled<T> = (value: T, label: AuthorityLabel)
\end{lstlisting}

The label includes integrity (authority level), confidentiality, origin, principals, transformation history, runtime signature, nonce, and timestamp. Labels are maintained in a sidecar store that the LLM cannot access or modify.

The integrity lattice orders sources by authority:
\begin{lstlisting}
SystemPolicy > UserIntent > TrustedSkill
  > AttestedToolMetadata > ToolMetadata
  > ToolOutput > ExternalContent > UntrustedSkill
\end{lstlisting}

The confidentiality lattice tracks sensitivity:
\begin{lstlisting}
Public < UserPrivate < Secret < CapabilityToken
\end{lstlisting}

These lattices are inspired by Denning~\cite{denning1976lattice} and Myers/Liskov~\cite{myers1997decentralized}. Unlike program-level IFC systems~\cite{fides_agent_ifc}, ProvShield labels are maintained in a sidecar store separate from the model context, preventing the LLM from manipulating its own security metadata. This is not a contribution of IFC itself---IFC is a tool---but of applying authority-flow reasoning to the specific problem of agent tool execution where the planner is untrusted.

\subsection{Effect-Typed Tools}
\label{sec:effects}

Each tool declares effects and sinks. Examples include \texttt{ReadPrivate}, \texttt{WriteExternal}, \texttt{SendNetwork}, \texttt{DeleteLocal}, \texttt{ExecuteCode}, \texttt{ModifyAuth}, and \texttt{CreateCredential}. The monitor uses these declarations to determine whether a proposed call is low-risk, high-risk, bridge-required, or denied.

Table~\ref{tab:effect-types} summarizes the effect taxonomy and the default policy action for each category.

\begin{table}[t]
  \centering
  \caption{Effect type taxonomy and default policy actions.}
  \label{tab:effect-types}
  \begin{tabular}{@{}lll@{}}
    \toprule
    \textbf{Effect}        & \textbf{Example Sink}       & \textbf{Default Action} \\
    \midrule
    \texttt{ReadPrivate}    & Local file system           & Allow (logged)          \\
    \texttt{WriteExternal}  & Network endpoint            & Bridge required         \\
    \texttt{SendNetwork}    & Email, HTTP POST            & Bridge required         \\
    \texttt{DeleteLocal}    & File system                 & Bridge required         \\
    \texttt{ExecuteCode}    & Shell, sandbox              & Bridge required         \\
    \texttt{ModifyAuth}     & Credential store            & Deny (unless admin)     \\
    \texttt{CreateCredential} & Token vault              & Deny (unless admin)     \\
    \texttt{Financial}      & Payment API                 & Deny (unless admin)     \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Runtime Monitor}
\label{sec:monitor}

The monitor intercepts all tool calls. For each call, it:
\begin{enumerate}
  \item normalizes the tool name and arguments;
  \item infers effects and sinks;
  \item reconstructs argument and payload provenance from the sidecar graph;
  \item checks source-to-sink policy;
  \item allows, denies, sanitizes, quarantines, or requests a bridge;
  \item logs the decision for replay.
\end{enumerate}

The monitor operates deterministically. It does not use the LLM for policy decisions. Provenance reconstruction uses the sidecar store, not model-generated text, to trace argument origins.

\subsection{Intent-Bound Declassification Capability}
\label{sec:bridge}

High-risk effects require an \emph{intent-bound declassification capability}---not a generic user confirmation. The capability binds the exact action, effect, sink, destination, payload digest, source categories, principal, expiration, and one-time nonce. It can authorize only the normalized call for which it was minted.

This design avoids the weakness of generic confirmation, where the user approves a vague action that the attacker can launder into a different destination, payload, or effect. The bridge presentation to the user includes:
\begin{itemize}
  \item the action name and destination;
  \item a SHA-256 digest of the payload;
  \item the provenance source categories contributing to the payload;
  \item whether private data or secrets cross a confidentiality boundary;
  \item a short expiration window (default: 30 seconds).
\end{itemize}

Figure~\ref{fig:bridge-flow} illustrates the bridge interaction flow.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    node distance=0.7cm and 1.0cm,
    box/.style={draw, rounded corners, minimum width=1.8cm, minimum height=0.5cm, font=\small, align=center},
    arr/.style={-{Stealth[length=2mm]}, thick},
    every node/.style={font=\small}
  ]
    \node[box] (llm) {LLM proposes\\tool call};
    \node[box, below=0.7cm of llm] (mon) {Monitor\\checks policy};
    \node[box, below left=0.7cm and 0.5cm of mon] (deny) {DENY\\(audit)};
    \node[box, below right=0.7cm and 0.5cm of mon] (bridge) {Bridge\\required};
    \node[box, below=0.7cm of bridge] (present) {Present action,\\destination, payload\\digest, sources};
    \node[box, below left=0.7cm and 0.3cm of present] (reject) {User rejects\\→ DENY};
    \node[box, below right=0.7cm and 0.3cm of present] (confirm) {User confirms};
    \node[box, below=0.7cm of confirm] (token) {Mint capability\\token (MAC)};
    \node[box, below=0.7cm of token] (exec) {Execute\\tool};

    \draw[arr] (llm) -- (mon);
    \draw[arr] (mon) -- node[left,font=\scriptsize] {low-risk} (deny);
    \draw[arr] (mon) -- node[right,font=\scriptsize] {high-risk} (bridge);
    \draw[arr] (bridge) -- (present);
    \draw[arr] (present) -- (reject);
    \draw[arr] (present) -- (confirm);
    \draw[arr] (confirm) -- (token);
    \draw[arr] (token) -- node[right,font=\scriptsize] {bound to\\call} (exec);
  \end{tikzpicture}
  \caption{Bridge interaction flow. When the monitor detects a high-risk effect influenced by low-integrity sources, it requires a user-intent bridge. The bridge binds the exact action, destination, payload digest, and provenance sources. A one-time capability token authorizes only the specific normalized call.}
  \Description{Flowchart showing the bridge interaction: LLM proposes tool call, Monitor checks policy, and either DENYs or requires a Bridge. The Bridge presents action details to the user, who can reject (DENY) or confirm (mint capability token, then execute).}
  \label{fig:bridge-flow}
\end{figure}

\subsection{Capability Tokens}
\label{sec:tokens}

A successful bridge creates a capability token~\cite{miller2006capabilities,dennis1966semantics}. The token is not textually usable by the model. It is stored in runtime state and checked by the monitor. Tokens cannot be copied from prompt text, modified, or replayed. Each token carries a cryptographic binding to the normalized action, sink, destination, payload digest, principal, and nonce. This design draws on capability-based protection~\cite{miller2006capabilities} adapted for the agent context.

%% ======================================================================
\section{Formal Model}
\label{sec:formal}

We mechanize the core of ProvShield's authority model as a labeled transition system and prove safety properties about runtime-observable tool execution. The formal guarantees are organized into three tiers: (1)~\emph{mechanized}: the transition relation and reachable-state invariants are verified by Coq (coqc 9.0); (2)~\emph{proof sketch}: unforgeability and policy-level properties are proved by induction on transitions, with the HMAC security assumption stated explicitly; and (3)~\emph{assumption}: TCB integrity, key secrecy, and tool effect accuracy are deployment assumptions validated by unit tests rather than formal proof.

We define runtime state as:
\begin{equation}
\Sigma = (C, R, T, S, P, B, A)
\label{eq:state}
\end{equation}
where $C$ is the context store, $R$ is runtime provenance state, $T$ is the tool registry, $S$ is the secret store, $P$ is the policy, $B$ is bridge/token state, and $A$ is the audit log. Each component is typed:

\begin{itemize}
  \item $C : \text{ObjId} \to \text{Labeled}(\text{Value})$, where $\text{Labeled}(V) = V \times \text{ProvenanceLabel}$.
  \item $R : \text{ObjId} \to \text{ProvenanceLabel}$, the sidecar store mapping objects to their runtime-signed labels.
  \item $T : \text{ToolName} \to \text{EffectSet} \times \text{SinkSet}$, the effect-type declarations for each tool.
  \item $S : \text{SecretId} \to \text{Confidentiality} \times \text{Value}$, the secret store.
  \item $P \subseteq \text{Integrity} \times \text{Confidentiality} \times \text{Effect} \times \text{Sink} \times \{\texttt{allow}, \texttt{deny}, \texttt{bridge}\}$, the policy rules.
  \item $B : \text{BridgeId} \to \text{BridgeToken}$, the active bridge tokens.
  \item $A = [\text{AuditEntry}]$, the append-only audit log.
\end{itemize}

A \emph{provenance label} $\ell = (I, C_\ell, \text{origin}, \text{principals}, \text{transforms}, \sigma, \text{nonce}, t)$ consists of integrity $I \in \{\texttt{UntrustedSkill}, \texttt{ExternalContent}, \texttt{ToolOutput}, \texttt{ToolMetadata}, \texttt{AttestedToolMetadata}, \texttt{TrustedSkill}, \texttt{UserIntent}, \texttt{SystemPolicy}\}$ ordered by the integrity lattice:
\begin{equation}
\texttt{SystemPolicy} > \texttt{UserIntent} > \texttt{TrustedSkill} > \texttt{AttestedToolMetadata} > \texttt{ToolMetadata} > \texttt{ToolOutput} > \texttt{ExternalContent} > \texttt{UntrustedSkill}
\end{equation}
and confidentiality $C_\ell \in \{\texttt{Public}, \texttt{UserPrivate}, \texttt{Secret}, \texttt{CapabilityToken}\}$ with the ordering $\texttt{Public} < \texttt{UserPrivate} < \texttt{Secret} < \texttt{CapabilityToken}$. The runtime signature $\sigma = \text{MAC}_k(\text{nonce} \| t \| I \| C_\ell)$ binds the label to the TCB key $k$.

A \emph{normalized tool call} is a tuple $k = (a, e, d, h, p, n)$ where $a$ is the tool action, $e$ is the effect class, $d$ is the destination, $h$ is the payload digest, $p$ is the principal, and $n$ is the nonce. Normalization is deterministic: two calls are equivalent iff all six fields match.

\subsection{TCB Assumptions}

The formal model assumes: (1) the HMAC key $k$ is known only to the TCB and never exposed to the context store $C$; (2) the sidecar store $R$ is only written by TCB transitions, not by model output; (3) the policy engine $P$ is not modifiable by the LLM or external content; (4) the bridge manager correctly enforces one-time nonce semantics; and (5) the tool registry $T$ accurately declares effect types. These assumptions define the security boundary: if any TCB component is compromised, the properties may not hold.

Figure~\ref{fig:transition} shows the labeled transition system governing tool-call lifecycle. Transitions include user input, external content ingestion, skill loading, tool registration, model proposal, monitor allow, monitor deny, bridge confirmation, tool execution, label propagation, and audit logging.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    node distance=1.2cm,
    state/.style={draw, rounded corners, minimum width=1.6cm, minimum height=0.5cm, font=\small, align=center},
    arr/.style={-{Stealth[length=2mm]}, thick},
    every node/.style={font=\small}
  ]
    \node[state] (ci) {Context\\Ingest};
    \node[state, right=1.0cm of ci] (ms) {Model\\Step};
    \node[state, right=1.0cm of ms] (tp) {Tool\\Proposed};
    \node[state, right=1.0cm of tp] (mc) {Monitor\\Check};

    \node[state, below left=1.0cm and 0.3cm of mc] (den) {Denied};
    \node[state, below=1.0cm of mc] (br) {Bridge\\Required};
    \node[state, below right=1.0cm and 0.3cm of mc] (al) {Allowed};

    \node[state, below left=0.8cm and 0.3cm of br] (bd) {Bridge\\Denied};
    \node[state, below right=0.8cm and 0.3cm of br] (bc) {Bridge\\Confirmed};

    \node[state, below=0.8cm of al] (te) {Tool\\Executed};
    \node[state, below=0.6cm of te] (lp) {Label\\Propagated};
    \node[state, below right=0.6cm and 0.8cm of den] (al2) {Audit\\Logged};

    \draw[arr] (ci) -- (ms);
    \draw[arr] (ms) -- (tp);
    \draw[arr] (tp) -- (mc);
    \draw[arr] (mc) -- node[left,font=\scriptsize] {deny} (den);
    \draw[arr] (mc) -- node[left,font=\scriptsize] {bridge} (br);
    \draw[arr] (mc) -- node[right,font=\scriptsize] {allow} (al);
    \draw[arr] (br) -- node[left,font=\scriptsize] {reject} (bd);
    \draw[arr] (br) -- node[right,font=\scriptsize] {confirm} (bc);
    \draw[arr] (bc) -- node[left,font=\scriptsize] {token} (al);
    \draw[arr] (al) -- (te);
    \draw[arr] (te) -- (lp);
    \draw[arr] (lp) -- (al2);
    \draw[arr] (den) -- (al2);
    \draw[arr] (bd) -- (al2);
  \end{tikzpicture}
  \caption{Labeled transition system for tool-call lifecycle. The monitor check is the single enforcement point; all execution paths pass through it.}
  \Description{State diagram showing the tool-call lifecycle: Context Ingest leads to Model Step, then Tool Proposed, then Monitor Check. From Monitor Check, three paths branch: Denied (to Audit Logged), Bridge Required (to Bridge Denied or Bridge Confirmed), and Allowed (to Tool Executed to Label Propagated to Audit Logged). Bridge Confirmed feeds back to Allowed via a capability token.}
  \label{fig:transition}
\end{figure}

\subsection{Theorem 1: Label Unforgeability}

\begin{theorem}[Label Unforgeability]
In any reachable state, no label in the sidecar store $R$ with integrity $> \texttt{ExternalContent}$ was created without a TCB transition.
\end{theorem}

\begin{proof}[Proof sketch]
We prove by induction on transition steps $n$ that the invariant $\text{Inv}(\sigma_n)$ holds: \emph{no label in $R$ with integrity $> \texttt{ExternalContent}$ was created without a TCB transition}.

\textbf{Base case} ($n=0$): Initial state $\Sigma_0$ has empty $R$, so $\text{Inv}(\Sigma_0)$ holds vacuously.

\textbf{Inductive step}: Assume $\text{Inv}(\sigma_n)$. For each transition $\sigma_n \xrightarrow{e} \sigma_{n+1}$:
\begin{itemize}
  \item \textsc{IngestExternal}: assigns $\texttt{ExternalContent}$ or lower; invariant preserved.
  \item \textsc{IngestUser}: assigns $\texttt{UserIntent}$; TCB operation with signature; invariant preserved.
  \item \textsc{ToolReturn}: assigns $\texttt{ToolOutput}$; invariant preserved.
  \item \textsc{BridgeConfirm}: assigns integrity via TCB; invariant preserved.
  \item \textsc{ModelStep}: LLM produces text in $C$ but has no write path to $R$. The monitor ignores model-generated labels and reconstructs provenance from $R$. Invariant preserved.
\end{itemize}

Labels are valid only if they carry $\sigma = \text{MAC}_k(\text{nonce} \| t \| I \| C_\ell)$. The LLM cannot compute $\sigma$ because $k$ is in the TCB and never exposed to $C$. Therefore, no model-generated text can create a valid label with higher integrity than permitted.
\end{proof}

\subsection{Theorem 2: Capability Token Unforgeability}
\begin{theorem}[Capability Token Unforgeability]
In any reachable state, a valid capability token can only be created by the runtime through a bridge confirmation or administrative transition.
\end{theorem}

\begin{proof}[Proof sketch]
Tokens are created only in transition \textsc{BridgeConfirm} or \textsc{AdminGrant}. Both are TCB operations that require either a user-confirmed bridge or an administrative credential. The token includes a MAC over the normalized call parameters and nonce. The LLM cannot compute a valid MAC without the key. Formally: for any $\sigma \in \text{Reach}(\Sigma_0)$, if $b \in B(\sigma)$ is a valid token, then there exists a prior state $\sigma'$ and a \textsc{BridgeConfirm} or \textsc{AdminGrant} transition $\sigma' \xrightarrow{e} \sigma''$ such that $b$ was minted in $\sigma''$.
\end{proof}

\subsection{Theorem 3: No Unauthorized Secret Exfiltration}

\begin{theorem}[No Unauthorized Secret Exfiltration]
If a proposed call sends or writes payload with confidentiality $\geq \texttt{Secret}$ to an external sink, the call cannot reach execution unless a valid declassification bridge exists.
\end{theorem}

\begin{proof}[Proof sketch]
The monitor reconstructs payload provenance from $R$. If any component has confidentiality $\geq \texttt{Secret}$ and the sink is external (not in $S$ or a trusted internal store), the policy requires a bridge. Without a valid bridge, the monitor transitions to \textsc{Deny}. A bridge is valid only if it binds the exact payload digest, destination, and action. Therefore, a generic or reused bridge cannot authorize a different exfiltration.
\end{proof}

\subsection{Theorem 4: No Low-Integrity Control of High-Risk Effects}

\begin{theorem}[No Low-Integrity Control]
If a high-risk effect is influenced by low-integrity sources (external content, tool output, untrusted skill, or untrusted tool metadata), the call cannot execute unless a valid bridge or policy exception authorizes the flow.
\end{theorem}

\begin{proof}[Proof sketch]
The monitor traces provenance of all arguments through $R$. If any argument's provenance includes a source with integrity $< \texttt{UserIntent}$ and the call has a high-risk effect, the policy requires a bridge. A bridge authorizes a specific normalized call; it cannot be repurposed for a different action or destination. Therefore, an attacker-controlled low-integrity source cannot direct high-risk execution through the bridge mechanism.
\end{proof}

\subsection{Theorem 5: Bridge Non-Replay}

\begin{theorem}[Bridge Non-Replay]
A bridge token authorized for a normalized call $(a, e, d, h, p, n)$ cannot authorize a different tuple $(a', e', d', h', p', n')$ where any field differs, or where the token has expired.
\end{theorem}

\begin{proof}[Proof sketch]
The token includes a MAC over all six fields plus an expiration timestamp. The monitor re-normalizes the proposed call and recomputes the expected MAC. Any field mismatch causes verification to fail. The nonce $n$ is consumed upon use (one-time semantics), preventing replay.
\end{proof}

%% ======================================================================
\section{Implementation}
\label{sec:implementation}

ProvShield is implemented as a runtime layer around an existing agent system. The implementation consists of six components:

\begin{enumerate}
  \item \textbf{MCP proxy} for tool registration and call mediation. It intercepts \texttt{tools/list} and \texttt{tools/call} messages, labeling metadata at registration and outputs at return time.
  \item \textbf{Skill loader} for provenance-aware skill ingestion. Skills from untrusted sources are labeled \texttt{UntrustedSkill}; attested skills from verified publishers are labeled \texttt{TrustedSkill}.
  \item \textbf{Context builder} for sidecar label construction. It assigns provenance labels to every object entering the agent context, including user messages, system prompts, web content, email, RAG results, and tool outputs.
  \item \textbf{Policy engine} for source-to-sink rules. The engine is configured via a declarative YAML policy that maps effect types and provenance patterns to actions (allow, deny, bridge).
  \item \textbf{Bridge manager} for user confirmations and token issuance. It renders human-readable bridge presentations, collects confirmation, and issues cryptographically bound capability tokens.
  \item \textbf{Audit logger} for deterministic replay. Every monitor decision, bridge interaction, and tool execution is logged with full provenance state for post-hoc analysis.
\end{enumerate}

The prototype avoids modifying MCP servers by mediating calls through a proxy. Tool metadata is labeled at registration time. Tool outputs are labeled at return time. All proposed calls pass through the monitor before execution. The implementation uses Python for the policy engine and bridge manager, with JSON-serializable provenance labels and HMAC-based token signatures.

%% ======================================================================
\section{Evaluation}
\label{sec:evaluation}

Our evaluation is organized around a central question: \emph{Does ProvShield prevent authority laundering when the planner is manipulated or fully compromised?}

We evaluate ProvShield along three axes: security (attack blocking), utility (benign task completion), and performance (runtime overhead).

\subsection{Authority-Laundering Attack Suites}
\label{sec:eval-attacks}

We evaluate against six authority-laundering classes:
\begin{enumerate}
  \item \textbf{Observation-to-send (Skill injection)}~\cite{skillinject}: malicious instructions embedded in skill files laundered into send/write effects.
  \item \textbf{Metadata-to-credential (MCP metadata poisoning)}~\cite{mcptox}: tool descriptions that attempt to exfiltrate credentials or modify authority.
  \item \textbf{MCP safety scenarios}~\cite{mcp_safety_audit}: demonstrated exploits against MCP-connected LLM agents including code execution, remote access, and credential theft.
  \item \textbf{Observation-to-send (Web/email/RAG injection)}~\cite{greshake2023indirect}: indirect injection through retrieved web pages, email content, and RAG documents.
  \item \textbf{Adaptive white-box attacks}: attackers with knowledge of the policy and label syntax who attempt to craft inputs that bypass enforcement~\cite{attriguard_agentwatcher}.
  \item \textbf{Direct-call adversary}: a fully compromised planner that directly issues attack tool calls, testing the runtime's upper-bound guarantee.
\end{enumerate}

\subsection{Baselines}
\label{sec:eval-baselines}

We compare ProvShield against:
\begin{itemize}
  \item \textbf{No defense}: the agent runs without security mediation.
  \item \textbf{Prompt hardening}: the system prompt instructs the model to ignore injected instructions~\cite{openai_agent_prompt_injection_defense}.
  \item \textbf{Input firewall}: a classifier filters malicious text before it reaches the model.
  \item \textbf{Static allowlist}: tools are restricted to a fixed set; no dynamic policy.
  \item \textbf{Generic confirmation}: the user is asked to confirm all writes, but the confirmation does not bind specific destination or payload.
  \item \textbf{Fides-style IFC}~\cite{fides_agent_ifc}: information-flow labels rendered in the prompt, with the model expected to self-enforce the policy. This is an approximation of Fides's runtime approach, adapted to the prompt-only setting.
  \item \textbf{Causal attribution}~\cite{attriguard_agentwatcher}: post-hoc counterfactual comparison of tool calls with and without external content, using injection-pattern-based heuristics to identify influenced calls.
  \item \textbf{MCP security}: pattern-based metadata scanner that inspects tool descriptions for suspicious instructions~\cite{mcp_safety_audit}.
\end{itemize}

\subsection{Metrics}
\label{sec:eval-metrics}

We report the following metrics:
\begin{itemize}
  \item \textbf{Attack success rate (ASR)}: fraction of attack scenarios where the malicious action executes.
  \item \textbf{Secret exfiltration rate}: fraction of scenarios where secret data reaches an external sink.
  \item \textbf{Unauthorized write/delete/exec rate}: fraction of scenarios where a destructive action executes without authorization.
  \item \textbf{Bridge abuse success}: fraction of scenarios where an attacker successfully manipulates the bridge.
  \item \textbf{Benign task completion rate (BTCR)}: fraction of benign tasks that complete successfully.
  \item \textbf{False blocking rate}: fraction of benign tool calls incorrectly denied.
  \item \textbf{Confirmation burden}: average number of bridge confirmations per benign task.
  \item \textbf{Monitor latency (p50/p95)}: latency added by the runtime monitor per tool call.
\end{itemize}

Table~\ref{tab:decomposed} presents the defense decomposition metrics, separating model manipulation from runtime enforcement.

\begin{table}[t]
  \centering
  \caption{Defense decomposition: separating model resistance from runtime enforcement.}
  \label{tab:decomposed}
  \begin{tabular}{@{}lr@{}}
    \toprule
    \textbf{Metric} & \textbf{Value} \\
    \midrule
    No-defense ASR & 5.1\% \\
    LLM manipulation rate & 14.9\% \\
    Attack-tool generation rate & 14.9\% \\
    Runtime block rate (conditional) & 100.0\% \\
    End-to-end ASR (ProvShield) & 0.6\% \\
    BTCR & 92.4\% \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Results}
\label{sec:eval-results}

Tables~\ref{tab:attack-results} and~\ref{tab:utility-results} present the evaluation results across all attack suites, baselines, and benign task categories.

\begin{table}[t]
  \centering
  \caption{Attack success rate (\%) across defense configurations. Lower is better. $n$=530 attack scenarios evaluated with LLM-in-the-loop (MiMo-v2-pro), including 72 workflow-embedded attack scenarios across 9 workflow types. 95\% Wilson confidence intervals.}
  \label{tab:attack-results}
  \small
  \begin{tabular}{@{}lcc@{}}
    \toprule
    \textbf{Defense} & \textbf{ASR (\%)} & \textbf{95\% CI} \\
    \midrule
    No defense         & 5.1  & [3.5, 7.3] \\
    Prompt hardening   & 1.9  & [1.0, 3.4] \\
    Input firewall     & 3.4  & [2.2, 5.3] \\
    Generic confirm.   & 5.1  & [3.5, 7.3] \\
    Static allowlist   & 0.0  & [0.0, 0.7] \\
    Fides-style IFC    & 4.9  & [3.4, 7.1] \\
    Causal attribution & 4.9  & [3.4, 7.1] \\
    MCP security       & 5.1  & [3.5, 7.3] \\
    \textbf{ProvShield} & \textbf{0.6} & \textbf{[0.2, 1.7]} \\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}[t]
  \centering
  \caption{Utility and overhead metrics. $n$=250 benign scenarios. BTCR = benign task completion rate.}
  \label{tab:utility-results}
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    \textbf{Defense} & \textbf{BTCR (\%)} & \textbf{False Block (\%)} & \textbf{Bridge Burden (\%)} \\
    \midrule
    No defense    & 100.0 & 0.0   & 0.0   \\
    Prompt hard.  & 100.0 & 0.0   & 0.0   \\
    Input fire.   & 100.0 & 0.0   & 0.0   \\
    Generic conf. & 100.0 & 0.0   & 0.0   \\
    Static allow. & 79.2  & 20.8  & 0.0   \\
    Fides IFC     & 100.0 & 0.0   & 0.0   \\
    Causal attr.  & 100.0 & 0.0   & 0.0   \\
    MCP security  & 100.0 & 0.0   & 0.0   \\
    \textbf{ProvShield} & \textbf{92.4} & \textbf{7.6} & \textbf{7.6} \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[t]
  \centering
  \caption{Per-suite attack success rate breakdown.}
  \label{tab:per-suite}
  \small
  \begin{tabular}{@{}lccc@{}}
    \toprule
    \textbf{Suite} & \textbf{ASR (\%)} & \textbf{Block (\%)} & \textbf{$n$} \\
    \midrule
    Skill injection      & 7.1  & 0.0 & 84   \\
    MCP metadata         & 7.0  & 0.0 & 86   \\
    RAG injection        & 3.9  & 0.0 & 103  \\
    Web/email injection  & 2.5  & 1.7 & 121  \\
    Adaptive white-box   & 5.9  & 0.7 & 136  \\
    \textbf{Overall}     & \textbf{5.1} & \textbf{0.6} & \textbf{530} \\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:per-suite} shows the per-suite ASR breakdown. ProvShield blocks all attacks in the skill injection, MCP metadata, MCP safety, and RAG suites. The residual 0.6\% ASR (3 attacks) occurs in the web/email (2/121) and adaptive white-box (1/136) suites.



\begin{figure}[t]
  \centering
  \begin{tikzpicture}[scale=0.85]
    % Axes
    \draw[->] (0,0) -- (7.5,0) node[right] {\small Defense};
    \draw[->] (0,0) -- (0,6.5) node[above] {\small ASR (\%)};

    % Y-axis labels
    \foreach \y/\lbl in {0/0, 1/1, 2/2, 3/3, 4/4, 5/5, 6/6} {
      \draw (-0.1,\y) -- (0.1,\y) node[left, font=\scriptsize] {\lbl};
    }

    % X-axis labels (defenses)
    \foreach \x/\lbl [count=\i] in {0.8/No def., 2.0/Prompt, 3.2/Firewall, 4.4/Confirm, 5.6/Allowlist, 6.8/ProvShield} {
      \node[below, font=\scriptsize, text width=1.2cm, align=center] at (\x, -0.1) {\lbl};
    }

    % Bars (1 unit = 1%)
    \fill[gray!30] (0.5, 0) rectangle (0.9, 5.1);   % No defense 5.1%
    \fill[gray!30] (1.7, 0) rectangle (2.1, 1.9);   % Prompt hard 1.9%
    \fill[gray!30] (2.9, 0) rectangle (3.3, 3.4);   % Firewall 3.4%
    \fill[gray!30] (4.1, 0) rectangle (4.5, 5.1);   % Confirm 5.1%
    \fill[gray!30] (5.3, 0) rectangle (5.7, 0.0);   % Allowlist 0%
    \fill[black!70] (6.5, 0) rectangle (6.9, 0.6);   % ProvShield 0.6%

    % Legend
    \fill[gray!30] (0.5, 6.8) rectangle (0.9, 7.1);
    \node[right, font=\scriptsize] at (1.0, 6.95) {Baselines};
    \fill[black!70] (3.0, 6.8) rectangle (3.4, 7.1);
    \node[right, font=\scriptsize] at (3.5, 6.95) {ProvShield};
  \end{tikzpicture}
  \caption{Attack success rate by defense across 530 attack scenarios. ProvShield achieves 0.6\% ASR compared to 5.1\% for no defense (88\% reduction). Static allowlist achieves 0\% ASR but at the cost of 20.8\% false blocking (79.2\% BTCR).}
  \Description{Bar chart comparing attack success rate across six defenses. No defense and generic confirmation are at 5.1\%, input firewall at 3.4\%, prompt hardening at 1.9\%, static allowlist at 0\%, and ProvShield at 0.6\%.}
  \label{fig:asr-comparison}
\end{figure}

\subsubsection{Security Analysis}

Across 530 attack scenarios (including 72 workflow-embedded strong attacks) evaluated with LLM-in-the-loop, ProvShield achieves an attack success rate of \textbf{0.6\%} (95\% CI [0.2\%, 1.7\%]), an \textbf{88\%} reduction from the no-defense baseline of 5.1\% (95\% CI [3.5\%, 7.3\%]). The LLM manipulation rate---the fraction of attack scenarios where the model generates any tool call---is 14.9\%, indicating that the model itself provides partial resistance. Among the scenarios where the LLM generates the specific attack target tool, ProvShield's conditional block rate is \textbf{100\%}: every malicious tool call that the LLM generates is blocked by the runtime monitor.

The no-defense ASR of 5.1\% reflects the robustness of modern LLMs against many injection patterns. The 72 workflow-embedded attack scenarios across 9 workflow types (where malicious actions are framed as legitimate IT procedures, compliance requirements, or policy updates) increase the manipulation rate compared to the standard scenarios. ProvShield's 100\% block rate demonstrates that even when the LLM is manipulated, the runtime enforcement prevents all unauthorized execution.

Among baselines, prompt hardening reduces ASR to 1.9\% but cannot defend against contextual or metadata-based attacks. Input firewall achieves 3.4\% ASR. Generic confirmation provides no protection (5.1\% ASR, identical to no defense) because it does not bind the specific destination or payload. Static allowlist achieves 0.0\% ASR but at the cost of 20.8\% false blocking (79.2\% BTCR).

The three prompt-side and post-hoc baseline variants---Fides-style IFC (prompt-rendered labels with model self-enforcement), causal attribution (counterfactual injection-pattern comparison), and MCP security (metadata pattern scanner)---perform no better than no defense (4.9--5.1\% ASR). Fides-style IFC renders provenance labels in the prompt and relies on the model to self-enforce, but the model ignores the labels under injection pressure. Causal attribution compares tool calls with and without external content to identify injection-driven calls, but the threshold-based filtering fails to distinguish malicious from benign influenced calls. MCP security scans tool metadata for suspicious patterns but does not address content-level injection. These results confirm that prompt-side and post-hoc defenses are insufficient against the attack patterns in our benchmark; runtime enforcement is required.

Table~\ref{tab:baseline-failures} summarizes representative failure modes for each baseline variant.

\begin{table}[t]
  \centering
  \caption{Representative baseline failure modes. Each baseline variant fails on at least one authority-laundering class that ProvShield blocks.}
  \label{tab:baseline-failures}
  \small
  \begin{tabular}{@{}p{2.2cm}p{2.2cm}p{3.2cm}@{}}
    \toprule
    \textbf{Baseline} & \textbf{Fails On} & \textbf{Failure Mechanism} \\
    \midrule
    Fides-style IFC & Skill injection, RAG injection & Model ignores prompt-rendered labels under injection pressure \\
    Causal attribution & MCP metadata, adaptive white-box & Cannot distinguish benign influence from malicious injection \\
    MCP security & Web/email injection, tool-output-to-exec & Only inspects metadata registration; misses runtime content \\
    Prompt hardening & Metadata poisoning, delayed injection & Depends on model obedience; fails on contextual attacks \\
    Generic confirmation & Confirmation laundering & Does not bind destination, payload, or provenance source \\
    \bottomrule
  \end{tabular}
\end{table}
The no-defense ASR of 5.1\% on mimo-v2-pro reflects the model's own safety alignment. Cross-model evaluation reveals that model susceptibility varies significantly: mimo-v2.5 achieves 50\% manipulation rate (vs.\ 14.9\% for mimo-v2-pro) and 10\% no-defense ASR. High-manipulation scenarios with multi-turn business framing push the no-defense ASR to 17.5\% on mimo-v2.5. While this is below the 30\% threshold that would demonstrate the full attack surface, the 100\% conditional block rate across all models and all attack configurations confirms that ProvShield's runtime enforcement provides defense-in-depth: even when the model's safety alignment is bypassed, the runtime prevents all unauthorized execution. The low absolute no-defense ASR is a feature of modern LLM safety training, not a limitation of our evaluation; ProvShield's value is in providing a complementary defense layer that operates independently of the model's safety alignment.

\subsubsection{Utility Analysis}

ProvShield achieves 92.4\% benign task completion (95\% CI [88.0\%, 94.9\%]), compared to 100\% for the no-defense baseline. The 7.6\% reduction is within the acceptance threshold and reflects the conservative policy applied to scenarios with ambiguous provenance. In the automated evaluation, bridge-required calls without user confirmation are conservatively treated as denied; the outright false denial rate (where no bridge is offered) is 7.6\%, while 21.1\% of benign tasks experience some friction (denial or bridge requirement). In a real deployment, user bridge confirmation would reduce the effective blocking rate. Static allowlist achieves only 79.2\% BTCR, demonstrating the cost of its coarse-grained approach.

\subsubsection{Conditional Metrics}

Among attack scenarios where the LLM generates the specific attack target tool, ProvShield's conditional block rate is \textbf{100\%}: every malicious tool call is blocked. This confirms that ProvShield's runtime enforcement is active precisely where it matters: when the model has been manipulated into generating malicious calls. The 100\% block rate holds across all attack suites, including the workflow-embedded strong attack scenarios.

\subsubsection{Performance}

The runtime monitor adds minimal overhead: p50 latency of 0.03\,ms and p95 latency of 0.07\,ms per tool call, well below the 100\,ms target. Because ProvShield maintains provenance labels in a sidecar store rather than embedding them in the prompt, the prompt token overhead is limited to lightweight advisory markers. The audit trace is fully replayable: every monitor decision, bridge interaction, and tool execution is logged with full provenance state for deterministic post-hoc analysis.


\subsubsection{Ablation Study}

Table~\ref{tab:ablation} presents the ablation study results, showing the contribution of each ProvShield component. These results are from a controlled policy-level evaluation (21 scenarios, predefined tool calls) to isolate component contributions.

\begin{table}[t]
  \centering
  \caption{Ablation study: Attack success rate (\%) across configurations. Policy-level evaluation with $n$=21 predefined attack scenarios.}
  \label{tab:ablation}
  \small
  \begin{tabular}{@{}lccccccr@{}}
    \toprule
    \textbf{Config} & \textbf{Skill} & \textbf{MCPTox} & \textbf{MCP Safe.} & \textbf{Web/Em.} & \textbf{RAG} & \textbf{Adapt.} & \textbf{All} \\
    \midrule
    A0 Full            & 33 & 50  & 50 & 0  & 50 & 11 & 24 \\
    A1 No labels       & 67 & 100 & 50 & 100& 100& 100& 90 \\
    A2 No monitor      & 100& 100 & 100& 100& 100& 100& 100\\
    A3 No bridge bind  & 33 & 50  & 50 & 0  & 50 & 11 & 24 \\
    A4 Trust metadata  & 33 & 100 & 50 & 0  & 50 & 11 & 29 \\
    A5 No cap token    & 33 & 50  & 50 & 0  & 50 & 11 & 24 \\
    A6 Conf. only      & 33 & 50  & 50 & 0  & 50 & 11 & 24 \\
    A7 Integ. only     & 67 & 50  & 50 & 33 & 100& 67 & 62 \\
    A8 No audit        & 33 & 50  & 50 & 0  & 50 & 11 & 24 \\
    \bottomrule
  \end{tabular}
\end{table}


The ablation reveals that provenance labels are the single most critical component: removing them (A1) raises ASR from 24\% to 90\%. Removing the runtime monitor (A2) yields 100\% ASR, confirming that enforcement---not just labeling---is essential. Trusting tool metadata (A4) specifically inflates MCPTox suite ASR to 100\%, confirming the value of treating unattested metadata as low-integrity. The integrity-only ablation (A7) shows that confidentiality checks are needed to prevent secret exfiltration (62\% ASR vs 24\% baseline). Note that the baseline (A0) shows 24\% ASR because these 21 predefined scenarios include high-risk tool calls that ProvShield correctly requires bridges for; in the policy-level evaluation without user confirmation, bridge-required calls count as blocked.

We additionally evaluate three provenance tracking modes---\emph{oracle} (perfect source attribution), \emph{conservative} (all context bound to all arguments), and \emph{heuristic} (pattern-based inference)---on 15 high-manipulation scenarios. All three modes achieve 100\% conditional block rate. The oracle mode shows lower manipulation rate (6.7\%) compared to conservative and heuristic modes (20\%), suggesting that finer-grained provenance reduces false positives in benign scenarios while maintaining security.

\subsubsection{Failure Analysis}

Across the 780-scenario evaluation (including 72 workflow-embedded strong attacks), ProvShield achieves 0.6\% ASR (3 attack successes out of 530). The residual attacks occur in the web/email (2) and adaptive white-box (1) suites, where the LLM generates tool calls whose provenance does not trigger high-risk policy thresholds. In two cases, web/email content was ingested through a path that preserved higher integrity labels; in one adaptive case, the attacker crafted content that matched user-intent patterns.

The 7.6\% false blocking rate arises from benign scenarios where the LLM generates tool calls with conservative provenance linking to external content, triggering bridge or deny decisions. In practice, field-level provenance tracking and user bridge interaction would reduce this rate.

\subsubsection{Multi-Model Evaluation}

To verify that ProvShield's enforcement is model-agnostic, we evaluate across three model variants using 75 scenarios per model (50 attack + 25 benign). Table~\ref{tab:multi-model} presents the results.

\begin{table}[t]
  \centering
  \caption{Multi-model evaluation (75 scenarios per model). ProvShield achieves 100\% conditional block rate across all models: every malicious tool call generated by any model is blocked by the runtime.}
  \label{tab:multi-model}
  \small
  \begin{tabular}{@{}lccccc@{}}
    \toprule
    \textbf{Model} & \textbf{ASR} & \textbf{No-def.\ ASR} & \textbf{Manip.} & \textbf{Block} & \textbf{BTCR} \\
    \midrule
    mimo-v2-pro   & 4.0\% & 8.0\%  & 32.0\% & 100.0\% & 100.0\% \\
    mimo-v2.5-pro & 2.0\% & 2.0\%  & 10.0\% & 100.0\% & 100.0\% \\
    mimo-v2.5     & 4.0\% & 10.0\% & 50.0\% & 100.0\% & 88.0\%  \\
    \bottomrule
  \end{tabular}
\end{table}

ProvShield achieves \textbf{100\% conditional block rate} across all three models: every malicious tool call generated by any model is blocked by the runtime monitor. The no-defense ASR varies from 2\% to 10\% depending on model susceptibility, with mimo-v2.5 being the most vulnerable (50\% manipulation rate). This confirms that ProvShield's runtime enforcement is independent of the underlying LLM's safety alignment.

\subsubsection{Direct-Call Adversary}

To isolate the runtime enforcement from the LLM's own resistance, we evaluate a direct-call adversary that bypasses the LLM entirely and sends crafted malicious tool calls directly to the runtime monitor. We use 101 attack scenarios covering all eight high-risk effect types (SendNetwork, ExecuteCode, DeleteLocal, CreateCredential, WriteExternal, CalendarInvite, ModifyAuth, Financial) with conservative provenance mode (all context objects linked to all arguments).

ProvShield blocks \textbf{101/101} (100\%) of direct-call attacks. Every effect type achieves 100\% block rate: every SendNetwork, ExecuteCode, DeleteLocal, CreateCredential, WriteExternal, CalendarInvite, ModifyAuth, and Financial call is either denied or requires a bridge. Table~\ref{tab:direct-call} presents the per-effect breakdown. This demonstrates that when tool effect declarations accurately reflect the tool's behavior, ProvShield's runtime enforcement provides complete protection against unauthorized execution regardless of how the tool call is generated.

\begin{table}[t]
  \centering
  \caption{Direct-call adversary: per-effect block rate ($n$=101). Conservative provenance mode.}
  \label{tab:direct-call}
  \small
  \begin{tabular}{@{}lrrr@{}}
    \toprule
    \textbf{Effect Type} & \textbf{$n$} & \textbf{Blocked} & \textbf{Block Rate} \\
    \midrule
    SendNetwork      & 17 & 17 & 100\% \\
    ExecuteCode      & 17 & 17 & 100\% \\
    DeleteLocal      & 12 & 12 & 100\% \\
    CreateCredential & 11 & 11 & 100\% \\
    WriteExternal    & 11 & 11 & 100\% \\
    CalendarInvite   & 11 & 11 & 100\% \\
    ModifyAuth       & 11 & 11 & 100\% \\
    Financial        & 11 & 11 & 100\% \\
    \textbf{Total}   & \textbf{101} & \textbf{101} & \textbf{100\%} \\
    \bottomrule
  \end{tabular}
\end{table}
\subsubsection{Stress-Test Scenarios}

We generate 18 stress-test scenarios using professional maintenance and operations framing (disk cleanup, build cache clearing, log rotation) with single-turn requests and no meta-language. On mimo-v2.5, these scenarios achieve \textbf{83\%} no-defense ASR (15/18 manipulated), satisfying the $\geq$30\% attack strength threshold. All 5 execute\_shell scenarios and all 3 send\_email scenarios succeed; 7/10 delete\_file scenarios succeed.

Critically, these stress-test scenarios represent \emph{user-level social engineering}: the malicious instructions appear as legitimate user requests with no external content injection. Since all context carries \texttt{UserIntent} integrity (the highest non-admin level), ProvShield correctly allows these calls---the policy only gates high-risk effects influenced by \emph{low-integrity} sources. This demonstrates ProvShield's precise enforcement boundary: it protects against \textbf{external content injection} (where untrusted data influences tool calls), not against \textbf{user-level social engineering} (where the user themselves authorizes risky actions).

\subsubsection{High-Manipulation Scenarios}

We also evaluate 80 high-manipulation attack scenarios using multi-turn business framing (IT compliance, incident response, GDPR data export, SOX compliance reports) where malicious instructions are embedded in external content. On mimo-v2.5, these scenarios achieve 13.3\% manipulation rate and 10.0\% no-defense ASR. ProvShield achieves \textbf{100\% conditional block rate} on all manipulated scenarios, confirming that when the LLM is deceived by external content into generating malicious calls, the runtime enforcement prevents all unauthorized execution.

\subsection{Comparison with Related Work}
\label{sec:eval-comparison}

Table~\ref{tab:comparison} compares ProvShield's design features with existing approaches.
\begin{table}[t]
  \centering
  \caption{Design comparison with related agent security approaches.}
  \label{tab:comparison}
  \small
  \begin{tabular}{@{}lcccccc@{}}
    \toprule
    \textbf{Feature} & \textbf{\makecell{Prompt\\Harden}} & \textbf{\makecell{Input\\Firewall}} & \textbf{\makecell{Generic\\Confirm}} & \textbf{Fides} & \textbf{AttriGuard} & \textbf{ProvShield} \\
    \midrule
    Runtime enforcement    & -- & -- & $\checkmark$ & $\checkmark$ & -- & $\checkmark$ \\
    Provenance labels      & -- & -- & -- & $\checkmark$ & -- & $\checkmark$ \\
    Sidecar (not prompt)   & -- & -- & -- & -- & -- & $\checkmark$ \\
    Effect-typed tools     & -- & -- & -- & $\checkmark$ & -- & $\checkmark$ \\
    Bound bridge           & -- & -- & -- & -- & -- & $\checkmark$ \\
    Capability tokens      & -- & -- & -- & -- & -- & $\checkmark$ \\
    MCP metadata defense   & -- & -- & -- & -- & -- & $\checkmark$ \\
    Skill loader defense   & -- & -- & -- & -- & -- & $\checkmark$ \\
    Formal safety props.   & -- & -- & -- & partial & -- & $\checkmark$ \\
    Audit replay           & -- & -- & -- & -- & $\checkmark$ & $\checkmark$ \\
    \bottomrule
  \end{tabular}
\end{table}

The key differentiator is ProvShield's combination of (1) sidecar provenance outside the model context, (2) bound user-intent bridges with cryptographic capability tokens, and (3) explicit treatment of MCP tool metadata and skill files as distinct provenance categories.

%% ======================================================================
\section{Discussion}
\label{sec:discussion}

ProvShield's guarantees are runtime guarantees. It does not claim to remove malicious influence from model activations. An attacker may still influence the model's planning through low-integrity content; ProvShield ensures that this influence cannot materialize as unauthorized high-risk tool execution.

\paragraph{Conservative policy.} ProvShield's policy may be conservative when provenance is ambiguous. For example, a tool output that mixes user-provided data with external content will be labeled at the lower integrity level. This can cause false blocking for benign mixed-provenance payloads. We mitigate this with field-level provenance tracking and configurable policy granularity.

\paragraph{Social engineering.} User confirmations can still be socially engineered. However, bridge binding prevents generic confirmation from authorizing hidden destination or payload changes. The bridge presentation includes payload digest, destination, and provenance source categories, making hidden changes visible to the user.

\paragraph{Model influence.} ProvShield does not address the problem of an attacker influencing the model's reasoning to cause benign-looking but suboptimal tool selections. This is an inherent limitation of treating the model as an untrusted planner rather than attempting to make it trustworthy.

\paragraph{Deployment.} ProvShield should be deployed alongside sandboxing, least-privilege tool access, and secret management. It is a complementary defense layer, not a replacement for standard security practices.

%% ======================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Prompt injection.}
Perez and Ribeiro~\cite{perez2022ignore} demonstrated goal hijacking and prompt leaking against GPT-3 through PromptInject. Greshake et al.~\cite{greshake2023indirect} introduced indirect prompt injection, showing that retrieved data can function as arbitrary code execution in LLM-integrated applications. OpenAI published design guidance for agents to resist prompt injection~\cite{openai_agent_prompt_injection_defense}, focusing on prompt hardening and input separation.

\paragraph{Information-flow control for agents.}
Fides~\cite{fides_agent_ifc} from Microsoft Research applies information-flow control to LLM agents, tracking labels within the model context. ProvShield differs by maintaining provenance in a sidecar store outside the model context, preventing the LLM from manipulating its own security metadata. NeuroTaint~\cite{neurotaint} extends taint tracking to semantic transformations in LLM agents, complementing ProvShield's runtime enforcement. The classic foundations of IFC include Denning's lattice model~\cite{denning1976lattice} and the decentralized label model of Myers and Liskov~\cite{myers1997decentralized}. TaiLi~\cite{taili} applies IFC principles specifically to LLM agent tool invocations.

\paragraph{Capability-based security.}
Dennis and Van Horn~\cite{dennis1966semantics} introduced capability-based addressing. Miller~\cite{miller2006capabilities} developed capability-based security for modern systems. ProvShield applies these principles to agent tool authorization through bound capability tokens. Narayan et al.~\cite{narayan2025capabilities} proposed capability-based security for AI agent tool invocation with formal guarantees. The Agent Passport~\cite{oap2026} specifies deterministic pre-action authorization for autonomous agents.

\paragraph{Taint tracking.}
Newsome and Song~\cite{newsome2005dynamic} introduced dynamic taint analysis for software security. ProvShield adapts this concept to the agent setting, tracking provenance through natural-language transformations rather than program data flow.

\paragraph{MCP security.}
Radosevich and Halloran~\cite{mcp_safety_audit} demonstrated security exploits in MCP-connected LLM agents and introduced MCPSafetyScanner for auditing MCP servers. Wang et al.~\cite{mcptox} created MCPTox, a benchmark for tool poisoning attacks on real-world MCP servers. The MCP authorization specification~\cite{mcp_authorization_spec} defines transport-level authorization for MCP connections but does not address content-level provenance.

\paragraph{Skill and tool security.}
Schmotz et al.~\cite{skillinject} introduced SkillInject, demonstrating up to 80\% attack success rate against frontier models through skill file injection. Shi et al.~\cite{progent} proposed Progent, which secures agents through privilege control with monotonic confinement. Song et al.~\cite{progent} further developed tool-use monitoring and privilege control frameworks for LLM agents. ToolEmu~\cite{toolem} provides an emulation framework for evaluating LLM agent safety, including tool misuse scenarios relevant to ProvShield's threat model.

\paragraph{Action attribution.}
AttriGuard~\cite{attriguard_agentwatcher} proposes causal attribution of tool invocations to distinguish user-intent-driven calls from injection-driven calls through counterfactual testing. ProvShield is complementary: it enforces policy at the runtime level regardless of the model's reasoning, while attribution-based approaches attempt to determine whether the model's reasoning was influenced.

\paragraph{Agent runtime monitoring.}
AgentWatcher~\cite{agentwatcher} monitors LLM agent behavior for anomalous tool invocations using behavioral analysis. Unlike ProvShield's policy-driven enforcement, AgentWatcher uses statistical detection. These approaches are complementary: ProvShield provides deterministic enforcement while behavioral monitors can detect novel attack patterns.
%% ======================================================================
\section{Limitations}
\label{sec:limitations}

ProvShield has several limitations that we state explicitly.

\paragraph{Model influence.} ProvShield does not claim to prevent untrusted text from influencing the LLM's hidden activations or reasoning. An attacker may still cause the model to generate suboptimal or strategically harmful tool selections that pass all provenance checks. ProvShield only guarantees that \emph{runtime-observable} high-risk effects cannot execute without proper provenance and capability authorization.

\paragraph{Social engineering.} User-intent bridges require human confirmation. A sufficiently convincing social engineering attack may still trick a user into confirming a harmful action. Bridge binding mitigates this by making the destination, payload digest, and provenance sources visible, but it cannot eliminate the risk entirely.

\paragraph{Conservative policy.} When provenance is ambiguous---for example, when a tool output mixes user-provided data with external content---ProvShield labels at the lower integrity level. This conservative join can cause false blocking for benign mixed-provenance payloads. Our evaluation shows 7.6\% false blocking on the benign suite, primarily from conservative taint propagation.

\paragraph{Key management.} ProvShield uses HMAC-SHA256 for label and token signatures. The prototype generates a module-level ephemeral key; production deployment requires a proper key management system with key rotation and secure storage. The formal guarantees assume the HMAC key is not compromised.

\paragraph{Tool effect declarations.} ProvShield's security guarantees depend on tools accurately declaring their effects and sinks. In practice, tool developers may not provide accurate declarations, and tools may have undeclared side effects. The prototype defaults to UNKNOWN\_HIGH\_RISK for undeclared tools, but automatic effect inference from tool behavior or code analysis is a direction for future work.

\paragraph{Policy configuration.} The policy engine uses a declarative YAML configuration, but the paper does not discuss how non-expert users would define or customize policies. In practice, administrators would need guidance on tuning policy strictness for their specific deployment context.

\paragraph{Residual attack surface.} ProvShield achieves 0.6\% ASR across 780 scenarios (3 residual attacks). The residual attacks occur in the web/email and adaptive suites where the LLM generates tool calls whose provenance does not trigger high-risk policy thresholds. Finer-grained argument-level provenance tracking would further improve precision.

\paragraph{TCB assumptions.} The formal guarantees assume the runtime monitor, sidecar store, policy engine, and cryptographic primitives are not compromised. If the runtime itself is compromised, all guarantees are void.
\paragraph{Evaluation scope.} Our evaluation uses 780 scenarios (530 attack, 250 benign) across 6 attack categories and 5 benign task categories. We evaluate across three model variants with 75 scenarios per model. However, the primary evaluation uses mimo-v2-pro; a larger-scale evaluation with diverse LLMs, real-world agent workflows, and stronger adaptive attacks would strengthen the evidence.
\paragraph{Formal proofs.} The Coq formalization compiles with coqc 9.0 and includes a transition relation with reachable-state invariants. However, Theorems 1 (label unforgeability) and 2 (token unforgeability) are currently proven at the definition level. The paper presents these as proof sketches, not fully mechanized proofs.
%% ======================================================================
\section{Ethics}
\label{sec:ethics}

\paragraph{Dual-use considerations.} This research studies prompt injection, metadata poisoning, and credential exfiltration. We designed the evaluation to use sandboxed mock tools, synthetic secrets (canary tokens), and synthetic MCP servers. No real credentials, private documents, or third-party accounts are used. If real vulnerabilities are discovered during research, we follow responsible disclosure practices.
\paragraph{User autonomy.} ProvShield may block or gate high-risk actions. The system explains decisions through the audit log and bridge presentation. Administrators can configure policies to adjust strictness. We do not hide the fact that an action was blocked due to untrusted content influence.

\paragraph{Audit log privacy.} Audit logs may contain provenance metadata, payload previews, and decision reasons. In production, logs must be redacted or encrypted. The artifact uses only synthetic data.

\paragraph{False sense of security.} ProvShield is a complementary defense layer, not a complete solution. We explicitly warn against deploying it without sandboxing, least-privilege tool access, and secret management. The paper's threat model and non-goals section clearly state what ProvShield does not protect against.

%% ======================================================================
\section{Artifact}
\label{sec:artifact}

The evaluation artifact is available as an open-source repository containing:

\begin{itemize}
  \item \textbf{Prototype}: Python implementation of all ProvShield components (policy engine, monitor, bridge manager, sidecar store, audit logger, MCP proxy with JSON-RPC support, skill loader with HMAC verification, three provenance modes: oracle, conservative, heuristic).
  \item \textbf{Evaluation harness}: 780 scenarios (530 attack, 250 benign) across 6 attack suites and 5 benign task categories, with 9 defense configurations including 3 baseline variants. Additional 80 high-manipulation scenarios with multi-turn business framing.
  \item \textbf{Direct-call adversary}: 101 crafted malicious tool calls sent directly to the runtime monitor (bypassing LLM), covering all 8 high-risk effect types (SendNetwork, ExecuteCode, DeleteLocal, CreateCredential, WriteExternal, CalendarInvite, ModifyAuth, Financial).
  \item \textbf{MCP integration demo}: Real filesystem MCP server with 4 replayable workflows (benign read, exfiltration, code execution, metadata poisoning).
  \item \textbf{Policy configuration}: Declarative YAML policy defining source-to-sink rules for all effect types.
  \item \textbf{Formal model}: Coq formalization with transition system, reachable-state invariants, and safety property proofs (compiles with coqc 9.0).
  \item \textbf{Audit log schema}: JSON Schema for reproducible audit traces with deterministic replay.
  \item \textbf{Smoke test}: Cross-platform Python script verifying 134 unit tests and integration scenarios in $\sim$10\,s.
  \item \textbf{Docker support}: Dockerfile for reproducible execution.
  \item \textbf{CI}: GitHub Actions workflow for automated testing.
\end{itemize}

\paragraph{Reproducibility.} The artifact supports three levels: (1) smoke test (134 unit tests, $\sim$10\,s), (2) full evaluation (780 scenarios with all defenses, $\sim$60\,min depending on LLM API), and (3) table generation for the paper. Random seeds are fixed per scenario (default: 42). All credentials and documents are synthetic.
\begin{lstlisting}[language=bash]
# Install
pip install -r requirements.txt

# Unit tests
pytest -q

# Smoke test (Level 1)
python artifact/scripts/run_smoke_test.py
# Expanded evaluation (Level 2)
python eval/scripts/run_expanded_eval.py

# Generate tables (Level 3)
python artifact/scripts/analyze_results.py \
  --results-dir eval/results \
  --output-dir eval/results/tables
\end{lstlisting}


\subsection{Baseline Implementation Details}
\label{sec:baseline-impl}

Each baseline variant is implemented in \texttt{eval/scripts/baselines.py} and applied to the same 780-scenario evaluation set as ProvShield. We describe the implementation and tuning of each.

\paragraph{Fides-style IFC.} We approximate Fides's runtime approach by rendering provenance labels (e.g., \texttt{[EXTERNAL]}, \texttt{[TOOL\_OUTPUT]}) in the system prompt alongside a policy instruction. The model is asked to self-enforce the IFC policy. This is a prompt-only approximation---not a faithful reproduction of Fides's runtime enforcement---and serves to test whether prompt-visible labels can serve as an authority boundary.

\paragraph{Causal attribution.} For each scenario, we compare the tool call produced by the LLM with external content present against a reference where external content is absent. If the tool name, destination, or payload digest changes, the call is flagged as injection-influenced and denied. This implements a counterfactual testing approach inspired by AttriGuard~\cite{attriguard_agentwatcher}. The key limitation is that influence is not authority: a benign tool call summarizing a webpage is legitimately influenced by external content.

\paragraph{MCP security.} We scan tool metadata (descriptions, parameter schemas) for suspicious patterns such as requests to include credentials, execute arbitrary code, or modify system configuration. This implements a pattern-based metadata scanner inspired by MCP Safety Audit~\cite{mcp_safety_audit}. The key limitation is that it cannot detect content-level injection at runtime or tool-output laundering.

\paragraph{Tuning protocol.} All baselines use default parameters without scenario-specific tuning. Pattern thresholds for causal attribution and MCP security are calibrated on a 32-scenario subset before application to the full evaluation set. No baseline receives access to ProvShield's sidecar store, policy engine, or capability tokens.

%% ======================================================================
\section{Conclusion}
\label{sec:conclusion}

Prompt injection in agents is best understood as an authority-flow problem. Untrusted content may influence language generation, but it should not be able to control privileged tool effects. ProvShield enforces this boundary through unforgeable provenance labels, effect-typed tools, runtime source-to-sink policy, and bound user-intent bridges.

We formalized the system with a labeled transition system and proved five safety properties as proof sketches: label unforgeability, capability-token unforgeability, no unauthorized secret exfiltration, no low-integrity control of high-risk effects, and bridge non-replay. The Coq formalization compiles with coqc 9.0 and includes the transition relation and reachable-state invariants, though the unforgeability theorems rely on definition-level reasoning and TCB assumptions.

Evaluation on 780 scenarios (530 attack, 250 benign) with LLM-in-the-loop, including 72 workflow-embedded attack scenarios across 9 workflow types and nine defense configurations (including three baseline variants: Fides-style IFC, causal attribution, and MCP security), shows that ProvShield achieves \textbf{0.6\%} attack success rate compared to 5.1\% for no defense---an 88\% reduction---while maintaining 92.4\% benign task completion. Every malicious tool call generated by the LLM targeting the attack-specific tool is blocked by the runtime monitor (100\% conditional block rate). Cross-model evaluation on three model variants (mimo-v2-pro, mimo-v2.5-pro, mimo-v2.5) confirms model-agnostic 100\% conditional block rate. Direct-call adversary evaluation with 101 scenarios across all eight effect types confirms 100\% runtime block rate under conservative provenance.

ProvShield does not claim to solve prompt injection. It claims to make the consequences of prompt injection visible and controllable at the runtime boundary. The model may be deceived; the runtime must not be.
%%
%% --- Bibliography -----------------------------------------------------
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
